---
title: "Uber Ride Analytics in NCR India"
format:
  revealjs:
    theme: simple
    transition: fade
    slide-number: true
    chalkboard: false
    center: false
    preview-links: true
editor: visual
---

```{r}
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(lubridate)
library(igraph)
library(visNetwork)
library(grid)

uber_data <- read_csv(
"https://raw.githubusercontent.com/tlwong1/Data-Visualization-Project-1/refs/heads/main/Project%20Data/ncr_ride_bookings.csv"
)

```

## Agenda

-   Policy use case & motivation

-   Data source & scope

-   Findings from visualizations

-   Interactive demo

-   Key takeaways & next steps

```{r}
#“I’ll briefly frame the policy motivation, walk through key findings on pricing and reliability, and then demonstrate an interactive tool for targeted decision-making.”
```

## Policy use case

Ride-hailing reliability and pricing transparency affect:

-   Two focus areas:

    -   Driver reliability and platform efficiency

    -   Fairness perceptions (especially for short urban trips)

```{r}
#“Ride-hailing platforms operate at large scale, where even minor inefficiencies can affect thousands of daily trips.”

#“From a policy and operations perspective, two issues matter most: pricing fairness for riders and reliability of service.”

#“Together, reliability and fairness perceptions shape customer trust and retention



#“This project asks whether short urban trips are systematically less cost-efficient, and whether cancellations cluster at specific times or locations.”
```

## Data source

-   **Dataset:** Uber Ride Analytics Dashboard dataset (Kaggle)

-   **Geography:** National Capital Region(NCR), India

-   **Unit of analysis:** individual bookings

-   **Key fields used:** Booking Status, Booking Value, Ride Distance, Time, Vehicle Type

```{r}
# “All figures are generated in R and can be reproduced from scratch using the attached code.”

#“All analysis is based on a publicly available Uber ride-booking dataset from Kaggle.”

#“The data covers individual bookings in India’s National Capital Region, a dense, high-volume urban market.”

#“Each observation represents a single ride, allowing us to analyze patterns at the individual level

#“The key variables used are booking status, Booking Value, ride distance, time of day, and vehicle type”
```

## Visualization 1: Price per Kilometer by Trip Distance

```{r}
#| echo: false
#| warning: false
#| message: false

# Clean data and calculate price per km
viz_data <- uber_data %>%
  mutate(
    booking_value = as.numeric(`Booking Value`),
    ride_distance = as.numeric(`Ride Distance`)
  ) %>%
  filter(
    !is.na(booking_value), 
    !is.na(ride_distance),
    booking_value > 50,      # Remove very low fares
    booking_value < 3000,    # Remove extreme outliers
    ride_distance > 0.5,     # Remove very short rides
    ride_distance < 50,      # Focus on typical rides
    `Booking Status` == "Completed"
  ) %>%
  mutate(
    price_per_km = booking_value / ride_distance,
    distance_category = cut(ride_distance, 
                           breaks = c(0, 5, 10, 15, 20, 50),
                           labels = c("0-5 km\n(Very Short)", 
                                    "5-10 km\n(Short)", 
                                    "10-15 km\n(Medium)", 
                                    "15-20 km\n(Long)",
                                    "20+ km\n(Very Long)"))
  )

# Create box plot showing price per km by distance category
ggplot(viz_data, aes(x = distance_category, y = price_per_km, fill = distance_category)) +
  geom_boxplot(outlier.alpha = 0.3, outlier.size = 0.5) +
  
  # Color scale
  scale_fill_viridis_d(option = "C", guide = "none") +
  
  # Labels
  labs(
    title = "Uber Price Per Kilometer by Trip Distance",
    subtitle = "Shorter trips cost significantly more per kilometer due to base fare minimums",
    x = "Trip Distance Category",
    y = "Price Per Kilometer (INR ₹/km)",
    caption = paste0("Data: Uber NCR Bookings (Kaggle). Analyzing ", 
                    format(nrow(viz_data), big.mark = ","), 
                    " completed rides.\nBox shows median and quartiles; dots are outliers.")
  ) +
  
  # Theme
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0, margin = margin(b = 5)),
    plot.subtitle = element_text(size = 10, color = "gray30", hjust = 0, margin = margin(b = 15)),
    plot.caption = element_text(size = 9, color = "gray50", hjust = 0, margin = margin(t = 10)),
    axis.title.x = element_text(face = "bold", size = 12, margin = margin(t = 10)),
    axis.title.y = element_text(face = "bold", size = 12, margin = margin(r = 10)),
    axis.text.x = element_text(size = 11),
    panel.grid.minor = element_blank(),
    plot.margin = margin(20, 20, 20, 20)
  ) +
  
  # Y-axis limits
  coord_cartesian(ylim = c(0, 150))
```

```{r}
#“This visualization examines pricing efficiency by calculating price per kilometer rather than total fare

#“Trips are grouped into distance categories to compare how cost efficiency changes with length.”

#“Very short trips under five kilometers are dramatically more expensive per kilometer than longer trips.”

#“The median price for these trips exceeds ₹100(rupees) per kilometer, compared to roughly ₹20 per kilometer for trips over 20 kilometers.”

#“This pattern likely occurs because base fare minimums dominate short trips, inflating per-kilometer costs.”

#“Short trips also show much greater variability, likely driven by surge pricing and time-of-day effects.”

#“As distance increases, pricing becomes both cheaper and more predictable.”
```

## Interpretation: Price vs Distance

-   Fares generally increase with distance, but not strictly linearly

-   Short trips show **greater pricing dispersion**

-   Distance alone does not explain what riders pay, especially for short urban trips

-   Perhaps base fare minimums / surges matter most for short urban rides

```{r}
#“Fares increase with distance, but not in a strictly linear way.”

#“Distance alone does not explain what riders pay, especially for short urban trips

#“Pricing efficiency improves substantially as trip length increases.”
```

## Visualization 2: Cancellation Rate by hour of day

```{r}
# Parse time and calculate cancellation rates by hour
hourly_data <- uber_data %>%
  mutate(
    hour = hour(hms(Time))
  ) %>%
  group_by(hour) %>%
  summarise(
    total_rides = n(),
    cancelled = sum(`Booking Status` %in% c("Cancelled by Customer", "Cancelled by Driver")),
    cancellation_rate = (cancelled / total_rides) * 100,
    .groups = "drop"
  ) %>%
  mutate(
    # Highlight morning and evening rush hours (7-9am, 5-7pm)
    is_rush_hour = hour %in% c(7, 8, 17, 18),
    hour_label = paste0(sprintf("%02d", hour), ":00")
  )

# Extract cancellation rates from the data
morning_val <- hourly_data$cancellation_rate[hourly_data$hour == 8]
evening_val <- hourly_data$cancellation_rate[hourly_data$hour == 18]

# Create dynamic annotation text
morning_text <- sprintf("Morning rush hour:\n%.1f%% cancellation rate", morning_val)
evening_text <- sprintf("Evening rush hour:\n%.1f%% cancellation rate", evening_val)

# Identify lowest and highest cancellation-rate points
min_point <- hourly_data[which.min(hourly_data$cancellation_rate), ]
max_point <- hourly_data[which.max(hourly_data$cancellation_rate), ]

# Average cancellation rate across all hours
avg_cancel <- mean(hourly_data$cancellation_rate, na.rm = TRUE)
avg_label  <- sprintf("Daily avg: %.1f%%", avg_cancel)

# Highlight the 14:00 (2pm) local peak
peak14_point <- hourly_data %>%
  filter(hour == 14)

peak14_text <- sprintf(
  "2pm local peak:\n%.1f%% at 14:00",
  peak14_point$cancellation_rate
)

# Create focused visualization
ggplot(hourly_data, aes(x = hour, y = cancellation_rate)) +
  
  # Background shading for rush periods
  annotate("rect", xmin = 7, xmax = 9, ymin = -Inf, ymax = Inf,
           alpha = 0.08, fill = "#E69F00") +
  annotate("rect", xmin = 17, xmax = 18, ymin = -Inf, ymax = Inf,
           alpha = 0.08, fill = "#E69F00") +
  
    # Average cancellation rate line (dashed)
  geom_hline(yintercept = avg_cancel, linetype = "dashed",
             linewidth = 0.9, color = "gray40") +

  
  # Base line
  geom_line(color = "gray60", linewidth = 1.2) +
  
  # Points
  geom_point(aes(color = is_rush_hour), size = 4) +
  
    # Highlight 2pm (14:00) local peak
  geom_point(
    data = peak14_point,
    aes(x = hour, y = cancellation_rate),
    color = "#009E73", size = 5, shape = 21, fill = "#009E73"
  ) +
  geom_text(
    data = peak14_point,
    aes(
      x = hour,
      y = cancellation_rate + 1.3,
      label = peak14_text
    ),
    color = "#009E73",
    fontface = "bold",
    size = 3.4,
    hjust = 0.5
  ) +

  
  # Highlight highest point (peak)
geom_point(data = max_point,
           aes(x = hour, y = cancellation_rate),
           color = "#D55E00", size = 5, shape = 21, fill = "#D55E00") +

geom_text(data = max_point,
          aes(x = hour, y = cancellation_rate + 1.5,
              label = sprintf("Highest point:\n%.1f%% at %02d:00",
                              cancellation_rate, hour)),
          color = "#D55E00", fontface = "bold", size = 3.5, hjust = 0.5) +

# Highlight lowest point (trough)
geom_point(data = min_point,
           aes(x = hour, y = cancellation_rate),
           color = "#0072B2", size = 5, shape = 21, fill = "#0072B2") +

geom_text(data = min_point,
          aes(x = hour, y = cancellation_rate - 1,
              label = sprintf("Lowest point:\n%.1f%% at %02d:00",
                              cancellation_rate, hour)),
          color = "#0072B2", fontface = "bold", size = 3.5, hjust = 0.5) +

  
  # Hour labels on highlighted points
  geom_text(
    data = subset(hourly_data, is_rush_hour),
    aes(label = sprintf("%02d:00", hour)),
    vjust = -1.2,
    size = 3,
    color = "#E69F00",
    fontface = "bold"
  ) +
  
  # MORNING ANNOTATION 
  annotate("curve",
           x = 9.7, xend = 8,
           y = morning_val + 4.2,
           yend = morning_val + 1,
           curvature = 0.3,
           arrow = arrow(length = unit(0.3, "cm")),
           color = "#E69F00", linewidth = 1) +
  annotate("text",
           x = 10, y = morning_val + 4.2,
           label = morning_text,
           hjust = 0,
           size = 3.8,
           color = "#E69F00",
           fontface = "bold") +
  
  # EVENING ANNOTATION
  annotate("curve",
           x = 15, xend = 18,
           y = evening_val - 2.8,
           yend = evening_val - 0.5,
           curvature = 0.3,
           arrow = arrow(length = unit(0.3, "cm")),
           color = "#E69F00", linewidth = 1) +
  annotate("text",
           x = 8.5, y = evening_val - 3,
           label = evening_text,
           hjust = 0,
           size = 3.8,
           color = "#E69F00",
           fontface = "bold") +
  
  # Colors
  scale_color_manual(
    values = c("FALSE" = "gray60", "TRUE" = "#E69F00"),
    guide = "none"
  ) +
  
  # Labels
  labs(
    title = "Uber Cancellation Rates Peak During Rush Hours",
    subtitle = str_wrap(
      "Cancellation rates are highest at 5am, suggesting driver shortage during peak commute times",
      width = 70
    ),
    x = "Hour of Day",
    y = "Cancellation Rate (%)",
    caption = "Data: Uber Ride Bookings, NCR India (Kaggle). Color palette: Okabe-Ito (CVD-safe)."
  ) +
  
  # Theme
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 15, hjust = 0),
    plot.subtitle = element_text(size = 11, color = "gray30", hjust = 0, margin = margin(b = 15)),
    axis.title = element_text(face = "bold", size = 12),
    axis.text = element_text(size = 11),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(color = "gray90", linewidth = 0.3)
  ) +
  
  scale_x_continuous(
    breaks = seq(0, 23, 3),
    labels = paste0(sprintf("%02d", seq(0, 23, 3)), ":00")
  ) +
  scale_y_continuous(
    breaks = seq(20, 30, 2),
    limits = c(20, 31)
  )
```

```{r}
#“This visualization shifts from pricing to reliability, focusing on when cancellations occur throughout the day.”

#“Cancellation rates remain relatively stable, fluctuating within a narrow five-percentage-point range.”

#“This stability indicates a generally resilient system rather than widespread breakdown.”

#“The highest cancellation rate occurs early in the morning around 5:00 AM, while the lowest occurs overnight.”

#“Morning and evening commute periods show a modest cancellations rate.”

#There is another peak at about 14:00 or 2pm
```

## Cancellation Rates Interpretation

-   **Cancellation rates are stable overall** (within \~5 percentage points)

<!-- -->

-   **Highest point occurs early morning** (low supply effect)

<!-- -->

-   **Commute windows have an average rate**

```{r}

#Cancellation rates are predicable

#Maybe the cancellation rate might do with supply issues at those times(5am, 2pm 14:00)
```

## Visualization 3: Peak Usage Hours

```{r}
# Filter for Auto rides and extract hour from Time
auto_hourly <- uber_data %>%
  filter(`Vehicle Type` == "Auto") %>%
  mutate(hour = as.numeric(substr(Time, 1, 2))) %>%
  count(hour) %>%
  arrange(hour)

# Create line/area chart showing peak hours
ggplot(auto_hourly, aes(x = hour, y = n)) +
  geom_area(fill = "skyblue", alpha = 0.6) +
  geom_line(color = "#2E86AB", linewidth = 1.2) +
  geom_point(color = "darkblue", size = 3) +
  scale_x_continuous(breaks = seq(0, 23, 2), 
                     labels = paste0(seq(0, 23, 2), ":00")) +
  scale_y_continuous(breaks = seq(0, 3000, 500)) +
  labs(
    title = "Peak Usage Hours for Auto Rides",
    subtitle = "Number of auto bookings by hour of day",
    x = "Hour of Day",
    y = "Number of Rides"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11, color = "gray30"),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

```{r}
#“This visualization shows when auto rides are most frequently requested.”

#“Demand is very low overnight, then rises sharply beginning around 4:00 AM.”

#“The first peak occurs in the late morning at 10, followed by a stronger peak around 6:00 PM.”

#“The evening peak exceeds the morning peak, indicating heavier return-commute 

#5am and 2pm(14:00) are right after the low points with 4am and 1pm(13:00) being the local minimum

#“The elevated cancellation rate align with the start of the demand peaks/spikes
```

## Interpretation: peak usage hours for auto

-   Demand follows a **predictable daily cycle**

<!-- -->

-   Two peaks: **late morning** and **stronger evening peak**

<!-- -->

-   Predictable timing enables **proactive driver supply planning**

```{r}
#“Auto ride demand follows a highly structured and predictable daily cycle.”

#“Demand spikes occur at nearly the same hours every day.”

#“This predictability creates an opportunity to proactively manage driver supply rather than reacting after problems emerge.”
```

## Visualization 4: Peak-hour comparison (weekday vs weekend)

```{r}
# 1) Auto rides by hour + weekday/weekend
auto_hourly <- uber_data %>%
  filter(`Vehicle Type` == "Auto") %>%
  mutate(
    hour = as.numeric(substr(Time, 1, 2)),
    day_type = if_else(wday(as.Date(Date), week_start = 1) %in% 1:5, "Weekday", "Weekend")
  ) %>%
  count(day_type, hour, name = "n") %>%
  arrange(day_type, hour)

# 2) Helper: peak + low points computed from auto_hourly
peak_points <- auto_hourly %>%
  mutate(
    window = case_when(
      hour >= 8  & hour <= 12 ~ "Morning Peak",
      hour >= 16 & hour <= 20 ~ "Evening Peak",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(window)) %>%
  group_by(day_type, window) %>%
  slice_max(n, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  transmute(
    day_type, window,
    x = hour, y = n,
    label = paste0(day_type, " ", window, ":\n", format(n, big.mark=","), " rides at ", sprintf("%02d:00", hour))
  )

low_points <- auto_hourly %>%
  filter(hour >= 10 & hour <= 16) %>%
  group_by(day_type) %>%
  slice_min(n, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  transmute(
    day_type,
    x = hour, y = n
  )

# 3) Offsets ONLY for evening-peak annotations
peak_points_annot <- peak_points %>%
  filter(window == "Evening Peak") %>%
  mutate(
    dx = if_else(day_type == "Weekday", 2.0, -2.0),
    dy = if_else(day_type == "Weekday", 520, 520),  # same baseline
    dy = if_else(day_type == "Weekend", dy + 600, dy), # nudge weekend up
    hjust_label = if_else(dx > 0, 0, 1)
  )

# 4) Δ line data between evening peaks 
eve_weekday <- peak_points_annot %>% filter(day_type == "Weekday") %>% slice(1)
eve_weekend <- peak_points_annot %>% filter(day_type == "Weekend") %>% slice(1)

diff_df <- tibble(
  hour = eve_weekday$x,
  y_min = min(eve_weekday$y, eve_weekend$y),
  y_max = max(eve_weekday$y, eve_weekend$y),
  label = paste0("Δ = ", format(abs(eve_weekday$y - eve_weekend$y), big.mark=","), " rides")
)

# 5) Plot
ggplot(auto_hourly, aes(x = hour, y = n, color = day_type, fill = day_type)) +
  geom_area(alpha = 0.25, position = "identity") +
  geom_line(linewidth = 1.2) +

  # Points 
  geom_point(data = peak_points, aes(x = x, y = y, color = day_type), size = 5) +
  geom_point(data = low_points,  aes(x = x, y = y, color = day_type), size = 5) +

  # Dotted Δ line at evening peak hour
  geom_segment(
  data = diff_df,
  inherit.aes = FALSE,
  aes(x = hour, xend = hour, y = y_min, yend = y_max),
  linetype = "dotted",
  linewidth = 1,
  color = "gray40"
) +
geom_label(
  data = diff_df,
  inherit.aes = FALSE,
  aes(x = hour + 3.5, y = (y_min + y_max)/2, label = label),
  fill = "white",
  color = "gray20",
  linewidth = 0.2,
  size = 3.4,
  fontface = "bold"
) +


  # Evening peak annotations only
  geom_curve(
    data = peak_points_annot,
    aes(x = x + dx, y = y + dy, xend = x, yend = y, color = day_type),
    curvature = 0.25,
    arrow = arrow(length = unit(0.25, "cm")),
    linewidth = 1
  ) +
  geom_text(
    data = peak_points_annot,
    aes(x = x + dx, y = y + dy, label = label, color = day_type, hjust = hjust_label),
    fontface = "bold",
    size = 3.3
  ) +

  coord_cartesian(clip = "off") +

  scale_x_continuous(
    breaks = seq(0, 23, 2),
    labels = paste0(sprintf("%02d", seq(0, 23, 2)), ":00")
  ) +
  scale_fill_manual(values = c("Weekday" = "#0072B2", "Weekend" = "#E69F00")) +
  scale_color_manual(values = c("Weekday" = "#0072B2", "Weekend" = "#E69F00")) +

  labs(
    title = "Peak Usage Hours for Auto Rides: Weekday vs Weekend",
    subtitle = "Weekday and weekend demand follow similar shapes, but at a different scale",
    x = "Hour of Day",
    y = "Number of Auto Rides",
    caption = "Data: Uber Ride Bookings, NCR India."
  ) +

  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11, color = "gray30"),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    legend.title = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.margin = margin(20, 30, 35, 30)
  )


```

```{r}
#“This visualization compares auto ride demand by hour of day for weekdays and weekends.”

#“Both curves follow a similar shape, indicating consistent daily timing patterns.”

#“However, weekday demand operates at a much higher scale.”

#“The weekday evening peak reaches 2,207 rides at 6:00 PM, compared to 857 rides on weekends.”

#“The dotted line highlights this gap of approximately 1,350 rides.”

#“This makes weekday evenings the most demand-intensive period for auto rides.”
```

## Interpretation: Weekday vs Weekend Demand

-   Peak-hour patterns are similar between weekdays and weekends

-   Suggests operational levers should be **calendar-aware**

-   Supports targeted policies rather than one-size-fits-all

```{r}
#“The main difference between weekdays and weekends is demand intensity, not timing.”

#“Work-related commuting compresses weekday demand into narrow windows.”

#“This concentration helps explain cancellation rate increases at the beginning of the rush hours.
```

## Visualization 5: High-Traffic Routes by Cancellation Rate

```{r}
# Calculate cancellation rates by route using Booking Status
route_analysis <- uber_data %>%
  filter(!is.na(`Pickup Location`), !is.na(`Drop Location`)) %>%
  group_by(pickup = `Pickup Location`, drop = `Drop Location`) %>%
  summarise(
    total_rides = n(),
    cancelled = sum(`Booking Status` %in% c("Cancelled by Customer", "Cancelled by Driver"), na.rm = TRUE),
    cancellation_rate = (cancelled / total_rides) * 100,
    .groups = "drop"
  ) %>%
  filter(total_rides >= 1,
         cancellation_rate >= 5) %>%
  arrange(desc(total_rides))

# Get TOP 50 busiest routes
problem_routes <- route_analysis %>%
  slice_head(n = 50)


# Calculate summary statistics for subtitle
total_routes <- nrow(problem_routes)
avg_cancel_rate <- mean(problem_routes$cancellation_rate)
worst_route <- problem_routes %>% 
  arrange(desc(cancellation_rate)) %>% 
  slice(1)

subtitle_text <- paste0(
  "Analyzing ", total_routes, " high-traffic routes with ≥5% total cancellation rates | ",
  "Average cancellation: ", round(avg_cancel_rate, 1), "% | ",
  "Worst route: ", worst_route$pickup, " → ", worst_route$drop, 
  " (", round(worst_route$cancellation_rate, 1), "%)"
)

# Create edges
edges_data <- problem_routes %>%
  rename(from = pickup, to = drop) %>%
  mutate(
    weight = cancellation_rate,
    label = paste0(round(cancellation_rate, 1), "%")
  )

# Get unique locations
all_locations <- unique(c(edges_data$from, edges_data$to))

# Get dropoff and pickup counts
dropoff_counts <- uber_data %>%
  filter(!is.na(`Drop Location`)) %>%
  count(`Drop Location`, name = "dropoff_total")

pickup_counts <- uber_data %>%
  filter(!is.na(`Pickup Location`)) %>%
  count(`Pickup Location`, name = "pickup_total")

# Create nodes
nodes_data <- data.frame(
  id = all_locations,
  label = all_locations
) %>%
  left_join(dropoff_counts, by = c("id" = "Drop Location")) %>%
  left_join(pickup_counts, by = c("id" = "Pickup Location")) %>%
  mutate(
    dropoff_total = replace_na(dropoff_total, 0),
    pickup_total = replace_na(pickup_total, 0),
    value = sqrt(dropoff_total) * 3,
    color = "#3498db",
    title = paste0("<b>", id, "</b><br>",
                   "Drop-offs: ", dropoff_total, "<br>",
                   "Pickups: ", pickup_total)
  )

# Calculate cancellations by type for each route
cancel_summary <- uber_data %>%
  filter(!is.na(`Pickup Location`), !is.na(`Drop Location`)) %>%
  group_by(pickup = `Pickup Location`, drop = `Drop Location`) %>%
  summarise(
    total_rides = n(),
    cancelled_by_customer = sum(`Booking Status` == "Cancelled by Customer", na.rm = TRUE),
    cancelled_by_driver = sum(`Booking Status` == "Cancelled by Driver", na.rm = TRUE),
    total_cancelled = cancelled_by_customer + cancelled_by_driver,
    cancellation_rate = (total_cancelled / total_rides) * 100,
    .groups = "drop"
  )


# Join without overwriting
edges_data_enhanced <- problem_routes %>%
  select(pickup, drop, total_rides, cancellation_rate) %>%
  left_join(
    cancel_summary %>% 
      select(pickup, drop, cancelled_by_customer, cancelled_by_driver, total_cancelled, cancellation_rate),
    by = c("pickup", "drop")
  ) %>%
  mutate(
    # Ensure no NAs and compute dominant side
    cancelled_by_customer = replace_na(cancelled_by_customer, 0),
    cancelled_by_driver = replace_na(cancelled_by_driver, 0),
    total_rides = replace_na(total_rides, 0),
    
    # Compute separate cancellation rates for each side
    rate_customer = (cancelled_by_customer / total_rides) * 100,
    rate_driver   = (cancelled_by_driver / total_rides) * 100,
    
    # Choose the higher rate to drive color
    dominant_rate = pmax(rate_customer, rate_driver, na.rm = TRUE),
    dominant_side = case_when(
      rate_customer > rate_driver ~ "Customer-driven",
      rate_driver > rate_customer ~ "Driver-driven",
      TRUE ~ "Equal"
    ),
    
    from = pickup,
    to = drop,
    label = paste0(round(dominant_rate, 1), "%"),
    title = paste0(
      "<b>Route:</b> ", from, " → ", to,
      "<br><b>Total Rides:</b> ", total_rides,
      "<br><b>Cancelled by Customer:</b> ", cancelled_by_customer,
      "<br><b>Cancelled by Driver:</b> ", cancelled_by_driver,
      "<br><b>Highest Cancellation Rate:</b> ", round(dominant_rate, 1), "%"
    ),
    
    # Color based on dominant (higher) cancellation rate
    color = case_when(
      dominant_rate >= 40 ~ "#8B0000",  # critical
      dominant_rate >= 30 ~ "#DC143C",  # severe
      dominant_rate >= 20 ~ "#FF6347",  # high
      dominant_rate >= 10 ~ "#FFA500",  # moderate
      TRUE ~ "#39FF14"                  # acceptable
    ),
    
    width = sqrt(total_rides) / 2
  ) %>%
  select(from, to, label, title, color, width, dominant_rate)

# Create combined title with stats
title_with_stats <- paste0(
  "<div style='text-align:center;'>",
  "<div style='font-size:20px; font-weight:bold; margin-bottom:8px;'>",
  "Uber NCR Problem Routes: Cancellation Rate Analysis",
  "</div>",
  "<div style='font-size:13px; color:#555; font-weight:normal;'>",
  "Analyzing ", total_routes, " high-traffic routes with ≥5% total cancellation rates | ",
  "Avg cancellation: ", round(avg_cancel_rate, 1), "% | ",
  "Worst: ", worst_route$pickup, " → ", worst_route$drop, 
  " (", round(worst_route$cancellation_rate, 1), "%)",
  "</div>",
  "</div>"
)

# Create network
visNetwork(nodes_data, edges_data_enhanced, height = "750px", width = "100%",
           main = title_with_stats) %>%
  visNodes(
    shape = "dot",
    color = list(background = "#3498db", border = "#2980b9"),
    font = list(size = 13, color = "#000000", bold = TRUE),
    borderWidth = 2
  ) %>%
  visEdges(
    arrows = list(to = list(enabled = TRUE, scaleFactor = 0.5)),
    smooth = list(type = "curvedCW", roundness = 0.2),
    font = list(size = 12, color = "#000000", align = "middle")
  ) %>%
  visPhysics(
    solver = "forceAtlas2Based",
    forceAtlas2Based = list(gravitationalConstant = -50),
    stabilization = list(iterations = 100)
  ) %>%
  visOptions(highlightNearest = list(enabled = TRUE, degree = 1, hover = TRUE)) %>%
  visInteraction(navigationButtons = TRUE)
```

```{r}
#“This visualization shifts the analysis from time-based patterns to geographic risk.”

#“Nodes represent locations, and edges represent high-traffic routes between them.”

#“Only routes with at least five percent cancellation rates are included.”

#“Node size reflects traffic(drop-off) volume, while edge color indicates cancellation severity.”

#“The average cancellation rate across these routes is approximately 25.7 percent.”

#“The most severe route exhibits a cancellation rate of 50 percent.”

#“Cancellations cluster in specific corridors rather than occurring randomly.”
```

## Interpretation: Route-Level Risk

-   High cancellation risk is **corridor-specific**

<!-- -->

-   High volume does not guarantee reliability

<!-- -->

-   Supports **route-specific interventions**

<!-- -->

-   Justifies **interactive drill-down**

```{r}
#“High cancellation rates are concentrated in a small number of corridors.”

#“High volume alone does not guarantee reliability.”

#“This supports route-specific interventions rather than city-wide policy changes.”

#“This is where interactive exploration becomes necessary.”
```

## Visualization 6: Shiny Interactive Tool

**Demo goal:** show how anyone can explore:

-   Differences in Vehicle Type

-   Cancellations by Driver or Customer only

-   Filter by minimum Cancellation rate

```{r}
#“This slide demonstrates the Shiny application that turns the network into a decision-support tool.”

#“The cancellation-rate threshold allows people to focus on the most severe routes.”

#“Filtering by cancellation type separates driver-driven from customer-driven failures.”

#“This distinction helps diagnose whether issues stem from supply constraints or rider experience.”

#“Hover details provide route-level metrics that support targeted intervention.”

#“The app enables precise prioritization rather than broad, inefficient responses.”
```

## Key takeaways

-   **Pricing efficiency:** short rides are costlier per km and less predictable

-   **Reliability:** cancellations fluctuate modestly but follow predictable timing

-   **Decision support:** interactive tools enable fast drill-down for operations

```{r}
#“Short trips are disproportionately more expensive per kilometer.”

#“Cancellations are stable overall with predictable stress windows.”

#“Route failures tend to cluster, making targeted intervention feasible.”
```

## Next steps

-   Model cancellation probability using hour + distance + other features

-   Add geography (pickup/drop) to locate repeated problem corridors

-   Improve measurement of surge / dynamic pricing if available

```{r}
#“Future work could incorporate driver supply data to directly measure supply–demand gaps.”

#like Modeling cancellation probability using some of the key variables like hour or distance

#“Time- and route-specific incentive pilots could be tested on the highest-risk corridors.

#5am and 2pm are some potential candidates.

#“Short-trip pricing experiments could evaluate whether reduced base fares improve efficiency and reliability.”

#“More formal evaluation methods could be used to quantify the impact of these targeted interventions in the future
```

## Thank you
